---
title: "EBS Pollock stock assessment"
subtitle: "Scaffold based on Alaska Groundfish Stock Assessment Guidelines (July 2024)"
author:
  - name: "Jim Ianelli"
  - name: "[Co-authors]"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

# Notes on this scaffold {.unnumbered}

This file is a **Tier 3 stock assessment scaffold** derived from the structure in:

- *Alaska Groundfish Stock Assessment Guidelines* (AFSC, July 2024)

Use this as a starting template for either:

- **Operational Full** assessments (complete detail), or
- **Operational Update** assessments (abbreviated sections with references to prior full assessment).

Where useful, this template includes prompts for using the `sporc` workflow in this repo.

# Introduction

## September Plan Team products

[Describe preliminary work, bridging runs, and proposed model changes reviewed in September.]

## November Plan Team products

[Describe final draft scope, products, and any constraints on post-submission changes.]

## December Council products

[Describe final outputs, presentations, and SIS reporting status.]

## Assessment product schedule

[State whether this is an operational full, update, or harvest projection year.]

## Unforeseen circumstances

[Document material timeline or data/process disruptions, if any.]

# Outline and format of stock assessments

[Optional short orientation text for reviewers.]

# Assessment definitions

[State assessment type and rationale.]

- Assessment type this year: **[Operational Full / Operational Update / Harvest Projection]**
- Tier: **3**
- Base model carried forward from: **[year/chapter]**

# Operational full and update guidelines

## Title page and authorship

[Include required authorship/affiliation/disclaimer text as needed by current SAFE conventions.]

## Executive summary

### Summary of changes in assessment inputs

[Bullet list of data/model/process updates since last accepted assessment.]

### Summary of results

[Concise summary of biomass, fishing mortality, reference points, OFL/ABC outcomes.]

### Responses to SSC and Plan Team comments

#### General stock assessment comments

[Response table or bullets.]

#### Comments specific to this assessment

[Response table or bullets.]

## Introduction

[Stock-specific context, geography, life history, fishery relevance.]

## Fishery and management history

[Management timeline, major changes, current regulations, catch history context.]

## Data

### Fishery data

[Catch, composition, effort, age/length availability, quality checks, key caveats.]

## Survey data

[Survey design, index generation, uncertainty, recent anomalies, inclusion/exclusion decisions.]

## Other data

[Tagging, maturity, environmental covariates, or other sources if used.]

## Analytic approach

[Describe model platform and configuration.]

### SPORC implementation notes

- Config file(s): `config/base.yml`, `config/scenarios.yml`
- Input building: `R/build_inputs.R` and/or `data-raw/build_ebs_pollock_data.R`
- Scenario runs: `analysis/run_scenarios.R`

```{r}
#| label: setup-sporc-results
#| include: false

suppressPackageStartupMessages({
  library(SPoRC)
  library(yaml)
  library(ggplot2)
  library(ggthemes)
  library(gt)
})

repo_root <- if (file.exists("../config/base.yml")) ".." else "."
repo_path <- function(...) {
  path <- file.path(...)
  if (grepl("^/", path)) path else file.path(repo_root, path)
}

source(repo_path("R", "build_inputs.R"))
source(repo_path("R", "extract_results.R"))

cfg <- yaml::read_yaml(repo_path("config", "base.yml"))
raw_data <- readRDS(repo_path(cfg$paths$data_rds))
base_fit <- readRDS(repo_path(cfg$paths$outputs_dir, "base.rds"))
input_list <- build_pollock_inputs(cfg, raw_data)

sporc_model_names <- "SPoRC base"

theme_assessment <- function() {
  ggthemes::theme_few(base_size = 11) +
    theme(
      legend.position = "bottom",
      strip.background = element_rect(fill = "grey92", color = "grey70"),
      panel.grid.minor = element_blank()
    )
}

array_df <- function(x, dim_names, year_dim = NULL, age_dim = NULL) {
  out <- as.data.frame.table(x, responseName = "value", stringsAsFactors = FALSE)
  names(out) <- c(dim_names, "value")
  for (nm in dim_names) {
    out[[nm]] <- as.integer(out[[nm]])
  }
  if (!is.null(year_dim)) {
    out$year <- raw_data$years[out[[year_dim]]]
  }
  if (!is.null(age_dim)) {
    out$age <- raw_data$ages[out[[age_dim]]]
  }
  out
}

bio_df <- rbind(
  transform(
    array_df(input_list$data$WAA, c("region", "year_index", "age_index", "sex"),
             year_dim = "year_index", age_dim = "age_index"),
    quantity = "Spawning weight at age"
  ),
  transform(
    array_df(input_list$data$MatAA, c("region", "year_index", "age_index", "sex"),
             year_dim = "year_index", age_dim = "age_index"),
    quantity = "Maturity at age"
  ),
  transform(
    array_df(base_fit$rep$natmort, c("region", "year_index", "age_index", "sex"),
             year_dim = "year_index", age_dim = "age_index"),
    quantity = "Natural mortality"
  )
)
bio_df <- bio_df[bio_df$year == max(raw_data$years), ]

make_use_df <- function(x, type, fleets) {
  out <- array_df(x, c("region", "year_index", "fleet"), year_dim = "year_index")
  out <- out[out$value != 0, ]
  out$type <- type
  out$fleet_name <- fleets[out$fleet]
  out
}

data_availability_df <- rbind(
  make_use_df(input_list$data$UseCatch, "Catch", "Fishery"),
  make_use_df(input_list$data$UseFishIdx, "Fishery index", "Fishery CPUE"),
  make_use_df(input_list$data$UseFishAgeComps, "Fishery age composition", "Fishery"),
  make_use_df(input_list$data$UseSrvIdx, "Survey index", c("BTS", "ATS", "AVO")),
  make_use_df(input_list$data$UseSrvAgeComps, "Survey age composition", c("BTS", "ATS", "AVO"))
)
data_availability_df$series <- paste(data_availability_df$type, data_availability_df$fleet_name)
sporc_ts <- extract_sporc_timeseries(base_fit, years = raw_data$years)

catch_df <- merge(
  transform(
    array_df(base_fit$rep$PredCatch, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    source = "Fitted"
  ),
  transform(
    array_df(input_list$data$ObsCatch, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    observed = value
  )[, c("region", "year_index", "fleet", "observed")],
  by = c("region", "year_index", "fleet")
)
catch_df$fleet_name <- "Fishery"

fish_idx_df <- merge(
  transform(
    array_df(base_fit$rep$PredFishIdx, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    fleet_name = "Fishery CPUE"
  ),
  transform(
    array_df(input_list$data$ObsFishIdx, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    observed = value
  )[, c("region", "year_index", "fleet", "year", "observed")],
  by = c("region", "year_index", "fleet", "year")
)
srv_idx_df <- merge(
  transform(
    array_df(base_fit$rep$PredSrvIdx, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    fleet_name = c("BTS", "ATS", "AVO")[fleet]
  ),
  transform(
    array_df(input_list$data$ObsSrvIdx, c("region", "year_index", "fleet"),
             year_dim = "year_index"),
    observed = value
  )[, c("region", "year_index", "fleet", "year", "observed")],
  by = c("region", "year_index", "fleet", "year")
)
index_df <- rbind(
  fish_idx_df[, c("region", "year", "fleet_name", "value", "observed")],
  srv_idx_df[, c("region", "year", "fleet_name", "value", "observed")]
)

selectivity_df <- rbind(
  transform(
    array_df(base_fit$rep$fish_sel, c("region", "year_index", "age_index", "sex", "fleet"),
             year_dim = "year_index", age_dim = "age_index"),
    fleet_name = "Fishery",
    fleet_type = "Fishery"
  ),
  transform(
    array_df(base_fit$rep$srv_sel, c("region", "year_index", "age_index", "sex", "fleet"),
             year_dim = "year_index", age_dim = "age_index"),
    fleet_name = c("BTS", "ATS", "AVO")[fleet],
    fleet_type = "Survey"
  )
)
selectivity_df <- selectivity_df[selectivity_df$year == max(raw_data$years), ]

nll_components <- c(
  Catch = sum(base_fit$rep$Catch_nLL, na.rm = TRUE),
  Fishery_index = sum(base_fit$rep$FishIdx_nLL, na.rm = TRUE),
  Survey_index = sum(base_fit$rep$SrvIdx_nLL, na.rm = TRUE),
  Fishery_age_composition = sum(base_fit$rep$FishAgeComps_nLL, na.rm = TRUE),
  Survey_age_composition = sum(base_fit$rep$SrvAgeComps_nLL, na.rm = TRUE),
  Recruitment = sum(base_fit$rep$Rec_nLL, na.rm = TRUE),
  Initial_recruitment = sum(base_fit$rep$Init_Rec_nLL, na.rm = TRUE),
  Fishing_mortality_penalty = sum(base_fit$rep$Fmort_nLL, na.rm = TRUE)
)
nll_df <- data.frame(
  component = gsub("_", " ", names(nll_components)),
  value = as.numeric(nll_components),
  row.names = NULL
)
nll_df <- nll_df[order(nll_df$value, decreasing = TRUE), ]

admb_path <- repo_path(cfg$paths$admb_results_rds)
admb_results <- if (file.exists(admb_path)) readRDS(admb_path) else NULL
admb_comparison <- if (!is.null(admb_results)) {
  compare_timeseries(sporc_ts, admb_results$timeseries)
} else {
  NULL
}
```

### Description of alternative models

[For full assessments: list alternatives, rationale, and selection logic.]

## Age structured model results

### Tables

[Insert required SAFE tables for Tier 3.]

```{r}
#| label: tbl-sporc-convergence
#| tbl-cap: "SPoRC base model convergence summary."

convergence_summary <- data.frame(
  model = sporc_model_names,
  objective = base_fit$optim$objective,
  convergence_code = base_fit$optim$convergence,
  convergence_message = base_fit$optim$message,
  pd_hessian = isTRUE(base_fit$sdrep$pdHess),
  max_abs_gradient = max(abs(base_fit$sdrep$gradient.fixed), na.rm = TRUE)
)

gt(convergence_summary) |>
  fmt_number(columns = c(objective, max_abs_gradient), decimals = 4) |>
  cols_label(
    model = "Model",
    objective = "Objective",
    convergence_code = "Convergence code",
    convergence_message = "Message",
    pd_hessian = "Positive-definite Hessian",
    max_abs_gradient = "Max. absolute gradient"
  )
```

```{r}
#| label: tbl-admb-comparison
#| tbl-cap: "Maximum absolute relative difference between the SPoRC base fit and the 2024 ADMB model for shared time series."

if (!is.null(admb_comparison)) {
  comparison_summary <- aggregate(
    abs(relative_difference) ~ quantity,
    data = admb_comparison,
    FUN = max,
    na.rm = TRUE
  )
  names(comparison_summary)[2] <- "max_abs_relative_difference"
  gt(comparison_summary) |>
    fmt_number(columns = max_abs_relative_difference, decimals = 4) |>
    cols_label(
      quantity = "Quantity",
      max_abs_relative_difference = "Max. absolute relative difference"
    )
} else {
  gt(data.frame(note = "ADMB comparison file was not available."))
}
```

### Figures

[Insert required SAFE figures for Tier 3.]

```{r}
#| label: fig-sporc-biological
#| fig-cap: "SPoRC biological inputs and biological process quantities for the base fit."
#| fig-width: 9
#| fig-height: 6
ggplot(bio_df, aes(x = age, y = value)) +
  geom_line(linewidth = 0.8, color = "#1b6ca8") +
  geom_point(size = 1.8, color = "#1b6ca8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(x = "Age", y = "Value") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-data-availability
#| fig-cap: "Data availability used in the SPoRC base fit."
#| fig-width: 9
#| fig-height: 5
ggplot(data_availability_df, aes(x = year, y = series)) +
  geom_point(aes(fill = type), shape = 21, size = 2.6, color = "grey20", alpha = 0.9) +
  labs(x = "Year", y = NULL, fill = "Data type") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-timeseries
#| fig-cap: "SPoRC base fit time series, including biomass, spawning biomass, recruitment, and fishing mortality."
#| fig-width: 9
#| fig-height: 6
ggplot(sporc_ts, aes(x = year, y = value)) +
  geom_line(linewidth = 0.8, color = "#1b6ca8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 2) +
  labs(x = "Year", y = "Value") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-catch-fits
#| fig-cap: "Observed and fitted catch for the SPoRC base fit."
#| fig-width: 9
#| fig-height: 5
ggplot(catch_df, aes(x = year)) +
  geom_line(aes(y = value), linewidth = 0.8, color = "#1b6ca8") +
  geom_point(aes(y = observed), size = 1.8, color = "grey15") +
  facet_wrap(~ fleet_name, scales = "free_y") +
  labs(x = "Year", y = "Catch") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-index-fits
#| fig-cap: "Observed and fitted survey and fishery indices for the SPoRC base fit."
#| fig-width: 9
#| fig-height: 6
ggplot(index_df, aes(x = year)) +
  geom_line(aes(y = value), linewidth = 0.8, color = "#1b6ca8") +
  geom_point(
    data = index_df[is.finite(index_df$observed), ],
    aes(y = observed),
    size = 1.8,
    color = "grey15"
  ) +
  facet_wrap(~ fleet_name, scales = "free_y", ncol = 2) +
  labs(x = "Year", y = "Index") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-selectivity
#| fig-cap: "Estimated fishery and survey selectivity for the SPoRC base fit."
#| fig-width: 9
#| fig-height: 6
ggplot(selectivity_df, aes(x = age, y = value, color = fleet_name)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ fleet_type, scales = "free_y") +
  labs(x = "Age", y = "Selectivity", color = "Fleet") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-likelihood
#| fig-cap: "Negative log-likelihood components for the SPoRC base fit."
#| fig-width: 9
#| fig-height: 6
ggplot(nll_df, aes(x = reorder(component, value), y = value)) +
  geom_col(fill = "#1b6ca8", width = 0.7) +
  coord_flip() +
  labs(x = NULL, y = "Negative log-likelihood") +
  theme_assessment()
```

```{r}
#| label: fig-sporc-admb-comparison
#| fig-cap: "Shared SPoRC and ADMB time series for the base-model bridge comparison."
#| fig-width: 9
#| fig-height: 6

if (!is.null(admb_results)) {
  admb_ts <- admb_results$timeseries
  admb_ts$model <- "ADMB 2024"
  bridge_ts <- rbind(
    sporc_ts[sporc_ts$quantity %in% unique(admb_ts$quantity),
             c("year", "quantity", "value", "model")],
    admb_ts[, c("year", "quantity", "value", "model")]
  )

  ggplot(bridge_ts, aes(x = year, y = value, color = model)) +
    geom_line(linewidth = 0.8) +
    facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
    labs(x = "Year", y = "Value", color = "Model") +
    theme_assessment()
}
```

### Evaluation of model(s) and associated uncertainty

[Diagnostics + uncertainty interpretation.]

### Sensitivity to model specification

[Parameter/process/data sensitivity cases and implications.]

### Convergence status and criteria

[Convergence checks, gradients, Hessian, warnings, remedies.]

### Likelihood profiles on key parameters

[Profiles and interpretation.]

### Retrospective analysis (within model)

[Retrospective setup/results and interpretation.]

### Historical retrospectives (between models)

[Cross-version retrospective comparisons, if applicable.]

## Projections and harvest recommendations

### Amendment 56 reference points

[Specify reference points used.]

### Specification of OFL and maximum permissible ABC

[Derivation and final values.]

### Standard harvest scenarios and projection methodology

[Projection assumptions, scenarios, and implementation details.]

## Risk table and ABC recommendation

[Risk table summary and recommended ABC rationale.]

## Area allocation of ABC

[Allocation method and resulting values.]

## Status determination

[Overfishing/overfished status determination text.]

## F\_limit

[If applicable under current guidance.]

## Ecosystem considerations

[Concise ecosystem context relevant to management advice.]

## Ecosystem effects on the stock

[Effects of ecosystem/state shifts on stock dynamics and uncertainty.]

## Fishery effects on the ecosystem

[Fishery/ecosystem interactions and management implications.]

## Data gaps and research priorities

[Prioritized list for next cycle(s).]

## Acknowledgements

[Contributors, reviewers, data providers, etc.]

## Literature cited

[Automatically generated from bibliography.]

## Auxiliary files

[List model/data/output artifacts and repository locations.]

## Tables

[Consolidated table section if required by chapter formatting.] 

## Figures

[Consolidated figure section if required by chapter formatting.]

## Appendices and working papers

[Include supplemental methods, diagnostics, and additional analyses.]

# Harvest projection guidelines (if this is a projection-only year)

## Title page and authorship

## Executive summary

### Description of updated catch

### Summary of results

### Harvest projection template

### Figures

# References

